Tutorial_4

### 2024-11-07 Tutorial_4.Rmd
tutorial 4 initial commit

---
title: "Tutorial_4"
author: "Sven Panis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

Tutorial_4 illustrates how to use functional programming to set up life tables, and plot the discrete-time hazard, survivor, and conditional accuracy functions, for the data of Experiment 1 of Panis and Schmidt (2016) when there are TWO independent variables (here: mask type and prime type).

# Load the libraries that we will be using.

```{r load-pkg}
pkg <- c("tidyverse", "patchwork")

@@

# Read in the raw data from Panis and Schmidt (2016).

```{r read-raw-data}
raw <- read_delim(file="Tutorial_1_descriptive_stats/data/DataExp1_6subjects.txt", delim="\t",col_names=T) # 9372 rows
head(raw)
tail(raw)
glimpse(raw)
```

# Do some wrangling.

```{r wrangle-raw-data}
# remove practice trials
data_wr <- raw %>% filter(practice==0) # 9240 experimental trials: 6 participants x 70 blocks x 22 trials-per-block

# remove unnecessary columns
data_wr <- data_wr %>% select(-c(ss,practice,rd,form,task,primeduration,
                                             blankduration,maskduration,targetduration,
                                             void,target_dir,corresp,resperr))
# look at trials where resp==0
data_wr %>% filter(resp==0) 
data_wr %>% filter(resp==0) %>% print(n=51)

# remove trials with no response info if rt < 800 (800 was the response deadline in each trial):
# the 46 trials where rt >= 800 and resp==0 are trials in which the subject did not respond; these should not be deleted because they will be treated as right-censored observations in EHA!
# the remaining 5 trials of subject 1 should be removed (for some technical reason the resp (1 or 2) was not recorded; perhaps (s)he pressed both buttons at the same time...)
sel = which(data_wr$resp==0 & data_wr$rt < 800)
data_wr <- data_wr[-c(sel),] # 9235 rows

summary(data_wr)
# vp = Versuchsperson or subject
# bl = block number
# tr = trial number within block
# mask_type: no mask (0), relevant mask (1), irrelevant mask (2), lines mask (3)
# prime_type: no prime or blank (0), consistent prime (1), inconsistent prime (2)
# resp = no button (0), button left (1), button right (2) 
# respac = accuracy
# rt = response time
# TrialNr = trial number within experiment
```

# Change variable names, and create factors.

```{r create-names-and-factors}
# change some variable names to required names (pid, trial, condition1, condition2, rt, and acc)
colnames(data_wr) <- c("pid","bl","tr","condition1", "condition2","resp","acc","rt","trial") 
# pid to indicate column with unique person IDs
# trial must be a unique trial number per subject
# condition1 must be a factor indicating the levels (1,2,...) and labels of the first independent variable (here: mask_type)
# condition2 must be a factor indicating the levels (1,2,...) and labels of the second independent variable (here: prime_type)
# rt to indicate column with response times (in ms)
# acc to indicate column with accuracies (1/0)

# create factors for conditions (mask type and prime type): the levels of condition1 are displayed across columns; the levels of condition2 are displayed within each plot.
data_wr <- data_wr %>% mutate(condition1 = condition1 + 1, # avoid level "0"
                              condition1 = factor(condition1, levels = c(1,2,3,4), labels = c("nomask","relevant","irrelevant","lines")),
                              condition2 = condition2 + 1, # avoid level "0"
                              condition2 = factor(condition2, levels = c(1,2,3), labels = c("blank","congruent","incongruent")))
summary(data_wr)
```

# Apply functional programming to set up life tables, and plot the discrete-time functions (h(t), S(t), and ca(t)).

Load the functions. Make sure you rename, and specify the same levels and labels for factors condition1 and condition2 in the function plot_eha_2IV as above.

```{r load-functions}
# function censor creates censored observations and discrete response times (drt), given a user-defined censoring time (timeout) and bin width in ms
censor <- function(df, timeout, bin_width){
  if(!(timeout %% bin_width == 0)){
    return("The censoring time must be a multiple of the bin width!")
  }
  if(timeout < 0 | bin_width < 0){
    return("Both timeout and bin_width must be larger than 0!")
  }
  df %>% mutate(right_censored = 0,
                rtc = ifelse(rt > timeout, timeout, rt) %>% round(digits=2), # censored response times
                right_censored = ifelse(rtc == timeout,1,right_censored), # 1 = right censored observation, 0 = observed rt
                drt = ceiling(rtc/bin_width), # drt = discrete response time
                cens_time = timeout, bin_width = bin_width) # save both user-defined parameters for plotting
}

# function ptb creates a person-trial-bin oriented data set
ptb <- function(df){
  df %>% uncount(weights = drt) %>% # create drt rows per trial
         group_by(trial) %>% 
         mutate(period = 1:n()) %>% # create time bin (or time period) ranks within each trial
         mutate(event = if_else(period == max(period) & right_censored == 0, 1, 0)) %>% # event = 1 indicates response occurrence
         ungroup()
}

# function setup_lt sets up a life table for each level 2 independent variables
setup_lt_2IV <- function(ptb){
  ptb %>% mutate(event = str_c("event", event)) %>%
          group_by(condition1,condition2,period) %>% 
          count(event) %>% 
          ungroup() %>% 
          pivot_wider(names_from = event,
                      values_from = n) %>% 
          mutate(event0 = ifelse(is.na(event0),0,event0), # replace NA with 0
                 event1 = ifelse(is.na(event1),0,event1),
                 risk_set = event0 + event1) %>% # define the risk set
          mutate(hazard = (event1 / risk_set) %>% round(digits = 3)) %>% # calculate hazard estimate
          mutate(se_haz = sqrt((hazard * (1 - hazard)) / risk_set) %>% round(digits = 4)) %>% # standard error for hazard
          group_by(condition1,condition2) %>%
          mutate(survival = (cumprod(1-hazard)) %>% round(digits = 4), # calculate survival estimate
                 term     = (cumsum(hazard / (risk_set * (1 - hazard)))) %>% round(digits = 7), # intermediate calculation
                 se_surv  = (survival * sqrt(term)) %>% round(digits = 5)  ) %>% # Greenwood's (1926) approximation
          ungroup() 
}

# function calc_ca calculates the conditional accuracies 
calc_ca_2IV <- function(df){
  df %>% filter(right_censored==0) %>%
         group_by(condition1,condition2,drt,cens_time,bin_width) %>%
         summarize(ca = mean(acc) %>% round(digits = 2),
                   n = n(),
                   .groups = 'drop') %>%
         ungroup() %>%
         mutate(period = drt,
                se_ca = sqrt((ca * (1-ca)) / n) %>% round(digits = 3)) %>%
         select(-drt)
}

# function join_lt_ca_2IV joins the conditional accuracies to the life tables when there are two independent variables
join_lt_ca_2IV <- function(df1,df2){df1 %>% left_join(df2, join_by(condition1,condition2,period))}

# function extract_median_2IV is used to extract the S(t).50 quantiles when there are two independent variables
extract_median_2IV <- function(df){
  above_pct50 <- df %>% 
      group_by(condition1,condition2) %>%
      filter(survival >= .5) %>% 
      slice(n()) # take last row
  below_pct50 <- df %>% 
      group_by(condition1,condition2) %>%
      filter(survival < .5) %>% 
      slice(1) # take first row
  # pull period above
  period_above <- pull(above_pct50, period)
  # pull survivor function values
  survival_above <- pull(above_pct50, survival)
  survival_below <- pull(below_pct50, survival)
  # estimate median by interpolation
  median_period <- period_above+((survival_above-.5)/(survival_above-survival_below))*((period_above+1)-period_above)
}

# function plot_eha_2IV plots the hazard, survivor, and ca(t) functions, when there are two independent variables
plot_eha_2IV <- function(df,subj,haz_yaxis){ 
  library(patchwork)
  cutoff <- df %>% pull(cens_time) %>% max(na.rm=T)
  binsize <- df %>% pull(bin_width) %>% max(na.rm=T)
  median_period <- extract_median_2IV(df)
  n_iv1 <- nlevels(df$condition1) 
  n_iv2 <- nlevels(df$condition2) 
  n_conditions <- n_iv1*n_iv2
  data_median <- c()
  for(i in 1:n_conditions){
    data_median <- append(data_median, c(median_period[i], median_period[i]))
  }
  # Create correctly named factors (here: prime and mask) and specify the same levels and labels as above
  data_medians <- tibble(period= data_median,
                         survival = rep(c(.5, 0),n_conditions),
                         prime =  factor(rep(rep(1:n_iv2, each=2),n_iv1),levels=c(1:3),labels=c("blank","congruent","incongruent")),
                         mask = factor(rep(1:n_iv1, each=2*n_iv2),levels=c(1:4),labels=c("nomask", "relevant", "irrelevant","lines")))

# plot the hazard functions
p1 <-df %>% 
  mutate(mask = condition1,       # Create correctly named factor
         prime = condition2) %>%  # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=hazard)) +
  geom_point(aes(y=hazard), size=1) + labs(color="Prime") +
  geom_linerange(aes(ymin=hazard-se_haz, ymax=hazard+se_haz), show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits = c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="h(t)", title = paste("Subject ", subj)) +
  theme(legend.background = element_rect(fill = "transparent"),
        panel.grid = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

# plot the survivor functions
p2 <-df %>%
  mutate(mask = condition1,      # Create correctly named factor
         prime = condition2) %>% # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=survival), show.legend = F) +
  geom_point(aes(y=survival), size=1, show.legend = F) +
  geom_linerange(aes(ymin=survival-se_surv, ymax=survival+se_surv), show.legend = F) +
  # add vertical lines to indicate the estimated median RTs with geom_path()
  geom_path(aes(x=period, y=survival),
            data = data_medians, 
            linetype = 3, show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits=c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="", y="S(t)") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

# plot the conditional accuracy functions
p3 <-df %>%
  mutate(mask = condition1,       # Create correctly named factor
         prime = condition2) %>%  # Create correctly named factor
  ggplot(aes(x=period, color=prime, group=prime)) + # Apply correctly named factor
  geom_line(aes(y=ca), show.legend = F) +
  geom_point(aes(y=ca), size=1, show.legend = F) +
  geom_linerange(aes(ymin=ca-se_ca, ymax=ca+se_ca), show.legend = F) +
  scale_x_continuous(breaks = c(1:(cutoff/binsize)), labels=c(1:(cutoff/binsize)*binsize),
                     limits=c(1,cutoff/binsize)) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x="Time bin t's endpoint (ms)", y="ca(t)") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~mask,nrow=1,ncol=4) # Apply correctly named factor

p1/p2/p3
}
```

Apply the functions

```{r apply-functions}
# create a data frame with two columns (pid and data); the latter contains a list of nested data frames
data_nested <- data_wr %>% group_nest(pid)
# apply the functions using map() and map2()
data_final <- data_nested %>% 
                     mutate(censored  = map(data, censor, 600, 40)) %>%   # ! user input: censoring time, and bin width
                     mutate(ptb_data  = map(censored, ptb)) %>%           
                     mutate(lifetable = map(ptb_data, setup_lt_2IV)) %>%      # create life tables without ca for 2 independent variables
                     mutate(condacc   = map(censored, calc_ca_2IV)) %>%       # calculate conditional accuracy for 2 independent variables
                     mutate(lifetable_ca = map2(lifetable, condacc, join_lt_ca_2IV)) %>%  # add ca to lifetables for 2 independent variables
                     mutate(plot      = map2(.x = lifetable_ca, .y = pid, plot_eha_2IV,1))  # create plots for 2 independent variables
```

# Extract info, and save plot for each subject

```{r extract-info, warning=F}
# view original data (column 2 in data_final) for subject 1:
pluck(data_final, 2, 1) %>% print(n=20)

# view person-trial-bin data set (column 4) for subject 1:
pluck(data_final,4,1) %>% print(n=30)

# view lifetable (column 7) for a specific combination of condition levels for subject 1:
pluck(data_final, 7, 1) %>% filter(condition1=="nomask" & condition2 == "blank")

# view plot (column 8) for subject 1:
pluck(data_final, 8, 1)

# view the median RTs in a tibble:
median_bin <- map(data_final$lifetable_ca, extract_median_2IV)
binsize    <- data_final$censored[[1]] %>% pull(bin_width) %>% max(na.rm=T)
median_rt  <- modify_depth(median_bin,1,~.x*binsize)
median_tb  <- map(median_rt, ~as_tibble(., rownames = "condition")) %>% list_rbind(names_to = "pid")
median_tb %>% print(n=6*12)
```

```{r save-plots, warning=F}
# save the plot for each subject
map2(paste0("Tutorial_4_descriptive_2IV/Plot_2IV_for_subject", data_final$pid, "_PanisSchmidt.png"), data_final$plot, ggsave, width = 12, height = 8, dpi = 600)
```

>>>>

---
title: "Tutorial_4"
author: "Rich"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
editor_options: 
  chunk_output_type: console
---

@@

Tutorial 4 shows one way to plan the design of a future experiment using EHA.

# 1. Load the libraries that we will be using #

```{r load-pkg, results='hide'}
pkg <- c("cmdstanr", "standist", "tidyverse", "RColorBrewer", "patchwork", 
         "brms", "tidybayes", "bayesplot", "future", "parallel", "VGAM", "faux")

@@

Set options.

```{r set-options, results='hide'}
options(brms.backend = "cmdstanr",
        mc.cores = parallel::detectCores(),
        future.fork.enable = TRUE,
        future.rng.onMisuse = "ignore") ## automatically set in RStudio

supportsMulticore()
detectCores()
```

```{r check-info, results='hide'}
# packageVersion("cmdstanr")
# devtools::session_info("rstan")
```

theme settings for ggplot

```{r plot-settings, results='hide'}
theme_set(
  theme_bw() +
    theme(text = element_text(size = 22, face = "bold"), 
          title = element_text(size = 22, face = "bold"),
          legend.position = "bottom")
)

## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

# 2. A brief outline of the workflow #

## 2.1 - Main background sources ##

The basic approach is based upon two main sources:

The basic structure and code follows the examples outlined by Solomon Kurz in his
'power' blog posts and Lisa Debruine's faux{} package.

For Solomon's blog posts, see here: [https://solomonkurz.netlify.app/tags/power/](https://solomonkurz.netlify.app/tags/power/)

For Lisa's faux package, see here: [https://debruine.github.io/faux/](https://debruine.github.io/faux/).
The vignettes provided by the faux package are particularly helpful and worth following before you take a deep dive in the below code.

And for some simpler worked examples of the general workflow (i.e., simpler than using EHA), then take a look at these examples by Rich, which show how you might simulate multi-level data for a range of factorial designs: [https://github.com/rich-ramsey/sim_demo](https://github.com/rich-ramsey/sim_demo)

## 2.2 - Basic workflow ##

The basic workflow is as follows:

1. Fit a regression model to an existing dataset.
2. Use the regression model parameters to simulate a one new dataset.
3. Create a function to iterate over many datatsets and vary parameters of interest (e.g., sample size, trial count, effect size).
4. Summarise the simulated data to estimate likely power or precision of the research design options.

## 2.3 - Exceptions to the normal workflow ##

The basic workflow above is a modified version of that recommended by Solomon and Lisa.
In a typical workflow, once the datasets are simulated, a regression model would be fit to each datatset.
Then, the parameters from each model would be summarised to make a judgment about the likely precision or power of the desing.
However, in the case of Model_4 in this tutorial, which took ~8 hours to build, we are not able to easily build 1000 models per variation in sample size and/or trial count, as that would take an unreasonably long time.
Instead, we choose to take a simpler approach and summarise the data per bin and condition, rather than the parameter estimates from the model.
Although this is not as satisfying, we feel it can still be a useful guide to set expectations. 
To further save time and computational resources, we also choose to use a much simpler model than Model_4, in order to outline the basic approach to planning via data simulation. 
The details on this simpler model structure are provided below.

# 3. Read in the data #

Here, we again use data form Panis and Schmidt (2016), but to make things simpler, we focus on only 6 timebins (timebins 4-9) and two conditions (con and inc).

## read in and wrangle the raw data ##

```{r load-data}
data <- read_csv("Tutorial_4_planning/data/tidyppp_Exp1_n6.csv") %>%
  ## no mask condition only, neutral prime removed, timebins 4-9 only
  filter(no_ma == 1, no_pr < 1, timebin %in% c(4:9)) %>% 
  ## remove unnecessary columns
  select(-c(bl, target_dir, corresp, resp, respac, rel_ma, irr_ma, ran_ma, no_ma)) %>%
  ## rename the error column
  rename(error = resperr) %>% 
  ## create a new variable and some factors
  mutate(prime = if_else(con_pr == 1, "con", "inc"),
         prime = factor(prime,
                        levels = c("con", "inc")),
         timebin = factor(timebin)) %>% 
  ## simplify the data by selecting a few variables only
  select(pid, timebin, prime, outcome)
head(data)
str(data)
```

data check

```{r check-conds}
data %>% 
  distinct(timebin, prime)
```

## create summary data ##

summary data

```{r calculate-summary-data}
data_pid <- data %>% 
  group_by(pid, timebin, prime) %>% 
  summarise(outcome = mean(outcome))
head(data_pid)

data_group <- data %>% 
  group_by(timebin, prime) %>% 
  summarise(outcome = mean(outcome))
head(data_group)
```

quick plot

```{r quick-summary-plot}
p3.1 <- ggplot(data_group, aes(x=timebin, y = outcome, colour = prime)) +
   geom_line(aes(group = prime)) + geom_point() +
   geom_jitter(data=data_pid, alpha = 0.5, width = 0.1, height = 0) +
   scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
   scale_x_discrete(labels = c(1:6)) +
   scale_fill_brewer(palette = "Dark2") +
   scale_colour_brewer(palette = "Dark2") +
   labs(title = "Panis & Schmidt (2016) (N=6)",
        x = "time bins")
p3.1
```

check the number of trials

```{r check-trials}
tally <- data %>% 
  group_by(pid, prime, timebin) %>%
  summarise(n=n(),
            sum=sum(outcome),
            .groups = "drop")
tally
```

# 4. Build an initial model #

In general, it can sometimes be a lot simpler to use an index coding approach to build regression models.

see this link for some background on an index coding approach:
https://bookdown.org/content/4857/the-many-variables-the-spurious-waffles.html#many-categories.

and then maybe specify the interaction term like this...
https://discourse.mc-stan.org/t/specifying-formulates-for-interaction-between-categorical-variables-with-the-index-coding-approach-in-brms/29449/3

And in the context of data simulation, index coding can be particularly useful, because it may be often easier and more intuitive to set condition means, rather than more complicated regression parameters, such as interaction terms.

So, the approach here is to build an initial model using an index coding approach, but with a much simpler dataset than has been used in prior tutorials.

Because we are building Bayesian regression models, which generate a posterior distribution per condition of interest in our index coding approach, we can simply calculate contrasts of interests and associated prevision or interval estimates (95% quantile intervals, for example) to address any questions we may have about comparisons between conditions, such as inc vs con in a particular timebin.

## formula ##

with varying intercept and slope for prime by pid

```{r model-formula}
formula = bf(outcome ~ 0 + timebin:prime +
               (0 + timebin:prime | pid))
```

## check the priors available ##

```{r priors}
get_prior(formula,
          data = data, family = bernoulli(link = "cloglog"))
```

## visualise priors ##

For an index coding model, priors will be set differently to how they were set in the earlier tutorials, because the earlier tutorials did not use an index coding approach.
Using an index coding approach, each condition has a prior for the mean of that condition.
And as we know from plot 3.1 above, the condition averages across the six time bins selected range from near zero on the hazard probability scale to 0.50. 
And in principle, the more general case could involve any hazard value from zero to 1.
Therefore, if we try to stick to a weakly informative approach to priors [see here for more details](https://github.com/stan-dev/stan/wiki/prior-choice-recommendations), then priors for the cloglog values should cover a fairly broad range.

For a look at values across the cloglog scale and how it compares to the logit scale, see [here](https://towardsdatascience.com/a-gentle-introduction-to-complementary-log-log-regression-8ac3c5c1cd83).

Ok, so let's take a look at some cloglog values at the lower and upper end of the probability scale.

```{r cloglog-vals}
cloglog_vals = clogloglink(c(0.01,0.99))
cloglog_vals

# -4.600149  1.527180
```

and now take a look at some distributions

```{r visualise-priors}
visualize("normal(0, 2)", "normal(0, 1)", "normal(0, 0.5)",
          xlim = c(-4, 4))
```

(0,2) looks like it covers all of our likely values, even if it is too broad.

## set priors ##

We set the prior for the mean as (normal(0,2)) and then we use some default recommendations from McElreath 2020 for the rest.

```{r set-priors}
priors <- c(
  set_prior("normal(0, 2)", class = "b"),
  set_prior("normal(0, 1)", class = "sd"),
  set_prior("lkj(2)", class = "cor")
)
```

## run the model ##

This model takes ~17 minutes to build on a 2020 MacBook Pro (2 GHz Quad-Core Intel Core i5).
For this tutorial, this chunk is skipped to save time. 
Change eval=TRUE in the code chunk to run this model.

```{r build-model, eval=FALSE}
plan(multicore)
bi <- brm(formula = formula,
        data = data, family = bernoulli(link = "cloglog"),
        prior = priors,
        iter = 2000, warmup = 1000, cores = 8, chains = 4,
        control = list(adapt_delta = 0.95),
        save_pars = save_pars(all=TRUE),
        seed = 123,
        init = 0.01,
        file = "Tutorial_4_planning/models/bi")
summary(bi)
```

The model built without any concerns, errors or warnings.

At this point, you would normally do the following:

- model checking by looking at the convergence of chains and model diagnostics. 
We'll skip this step for now, just to move on to the main focus of planning and data simulation.

- Summarise and visualise the parameter estimates and/or posterior predictions.
Below, we quickly summarise fixed effects from the model in cloglog values, and we also convert them to hazard values, which tend to be far easier to interpret.

## visualise fixed effect parameters from model bi ##

read in the existing model object (if you did not build the model yourself)

```{r load-model-object}
bi <- readRDS("Tutorial_4_planning/models/bi.rds")
summary(bi)
```

wrangle the posterior distribution

```{r wrangle-posterior}
## take the posterior draws
post <- as_draws_df(bi) %>% 
  select(-lp__) %>% 
  as_tibble()

## create a summary, and use medians with robust = TRUE
post_summary <- posterior_summary(bi, robust = TRUE)

## just look at the fixed effects
post_qi_b <- post %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything()) %>%
  group_by(name) %>% 
  median_qi(value)
head(post_qi_b)
```

a quick plot using posterior samples and tidybayes

```{r plot-fixed}
tidy_fixed <- post %>% 
  select(starts_with("b_"), .chain, .iteration, .draw) %>% # select and rename in simpler labels.
  rename(chain=.chain, iter=.iteration, draw=.draw) %>% 
  pivot_longer(-c(chain, draw, iter)) %>% # move from wide to long
  mutate(key = factor(name, levels=unique(name)),
         bin = rep(1:6, length.out = 48000),
         prime = rep(c("con", "inc"), each = 6, length.out = 48000),
         bin=factor(bin),
         prime=factor(prime))
head(tidy_fixed)
tail(tidy_fixed)

check.labels <- tidy_fixed %>% 
  distinct(key, bin, prime) 
check.labels

# plot
p4.1 <- ggplot(tidy_fixed, aes(x = bin, y = value, fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "Model bi cloglog",
       x = "timebin", y = "cloglog") +
  scale_fill_brewer(palette = "Dark2") 
p4.1

ggsave ("Tutorial_4_planning/figures/index_cloglog.jpeg",
        width = 6, height = 4)
```

calculate hazards per condition per bin in the posterior dist

```{r calculate-bi-hazard}
tidy_haz <- tidy_fixed %>%
  select(chain, iter, draw, bin, prime, value) %>% 
  mutate(hazard = exp(value))
head(tidy_haz)
```

plot hazards

```{r plot-bi-hazards}
p4.2 <- ggplot(tidy_haz, aes(x = bin, y = hazard,
                                    fill=prime)) +  
  stat_halfeye(alpha=0.7) +
  labs(title = "hazard estimates",
       x = "timebin", y = "hazard") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) 
p4.2

ggsave ("Tutorial_4_planning/figures/index_hazard.jpeg",
        width = 6, height = 4)
```

warning about removing rows.

# 5. Simulate a single dataset #

Note: For multi-level data, data simulation can be quite a convoluted process, and we therefore strongly recommend reading the power blogs by Solomon Kurz and working through the faux package vignettes, before getting stuck into the below code.

read in a prior model, if not already loaded.

```{r load-model-object-sim, eval=FALSE}
bi <- readRDS("Tutorial_4_planning/models/bi.rds")
summary(bi)
```

Define some parameters

```{r define-sim-parameters}
# define parameters
# specify some features of the design
subj_n = 10  # number of subjects
rep_n = 200 # number of trial repeats 
cond_n = 2 # number of conditions
bin_n = 6 # number of time bins

## set fixed effects
## as a reminder
fixef <- as_tibble(fixef(bi), rownames = "term") %>%
  mutate(across(where(is.double), \(x) round(x, 2)))
fixef

## b1 to b6 = timebin 1 to 6.
## c = con
b1c = -4.66
b2c = -3.24
b3c = -2.32
b4c = -1.49
b5c = -1.21
b6c = -0.60

## i = inc
b1i = -4.37
b2i = -3.21
b3i = -2.37
b4i = -2.27
b5i = -3.04
b6i = -2.45

## set varying effects by pid
## as a reminder
varcor <- VarCorr(bi)
glimpse(varcor)
str(varcor)

## extract sd
sd <- as_tibble(varcor$pid$sd, rownames = "term") %>%
  mutate(across(where(is.double), \(x) round(x, 2)))
sd

u1c_sd = 1.06   # 
u2c_sd = 1.34   # 
u3c_sd = 0.60   # 
u4c_sd = 0.47   # 
u5c_sd = 0.58   #
u6c_sd = 0.72   #

u1i_sd = 1.17   # 
u2i_sd = 1.05   # 
u3i_sd = 0.91   # 
u4i_sd = 1.47   # 
u5i_sd = 1.34   #
u6i_sd = 0.62   # 
```

make a correlation matrix

```{r define-cor-mat}
# varcor <- VarCorr(b3c_out)
# varcor
# str(varcor)

## take each one
sd <- varcor$pid$sd
cor <- varcor$pid$cor
cov <- varcor$pid$cov

## make it tidy
cor_mat_tidy <- as_tibble(cor, rownames = "term") %>% ## 
  rename_with(~str_replace_all(.x, '\\.', '_')) %>% 
  pivot_longer(-term,
               values_to = "value",
               names_to = c("parameter", "term2"),
               names_pattern = "([a-zA-Z_\\d?]+)_([a-zA-Z:\\d?]+)") %>% ## pivot longer and use names_pattern to select the parts of the column names to separate on. the above is regex for each bit between the separator ("-"). e.g., ([a-zA-Z]+\\d+). this looks for letters and digits before the first "_". This (\\d?) looks for the possibility of digits.
  ## the rest of this code just turns the variables into what we want
  filter(parameter == "Estimate") %>%
  select(-parameter) %>% 
  pivot_wider(names_from = "term2",
              values_from = "value") %>% 
  select(-term)
head(cor_mat_tidy)
str(cor_mat_tidy)

## check with the model summary
# summary(bi)

## remove names and make a matrix as faux{} likes it like that
## full cor mat
cor_mat <- unname(as.matrix(cor_mat_tidy))
cor_mat
```

setup the data structure

```{r setup-sim-d1}
# make it reproducible
set.seed(1)

d1 <- add_random(subj = subj_n, rep = rep_n) %>%
  add_within("subj", condition = c("cond1", "cond2")) %>%
  add_contrast("condition", "treatment", add_cols = TRUE, 
               colnames = c("cond")) %>%
  add_within("rep", bin = 1:bin_n) %>%
  ### create new conditions here that respect the twelve levels b1c, b1i etc.
  mutate(con = if_else(condition == "cond1", 1, 0),
         inc = if_else(condition == "cond2", 1, 0)) %>% 
  mutate(bin1 = if_else(bin == 1, 1, 0),
         bin2 = if_else(bin == 2, 1, 0),
         bin3 = if_else(bin == 3, 1, 0),
         bin4 = if_else(bin == 4, 1, 0),
         bin5 = if_else(bin == 5, 1, 0),
         bin6 = if_else(bin == 6, 1, 0)) %>% 
  # add random effects 
  add_ranef("subj", u1c = u1c_sd, u2c = u2c_sd, u3c = u3c_sd, u4c = u4c_sd,
            u5c = u5c_sd, u6c = u6c_sd,
            u1i = u1i_sd, u2i = u2i_sd, u3i = u3i_sd, u4i = u4i_sd,
            u5i = u5i_sd, u6i = u6i_sd, 
            .cors = cor_mat) %>% 
  # calculate logit
  mutate(cloglog = (b1c + u1c) * con * bin1 + (b2c + u2c) * con * bin2 + 
           (b3c + u3c) * con * bin3 + (b4c + u4c) * con * bin4  + 
           (b5c + u5c) * con * bin5 + (b6c + u6c) * con * bin6 +
           (b1i + u1i) * inc * bin1 + (b2i + u2i) * inc * bin2 + 
           (b3i + u3i) * inc * bin3 + (b4i + u4i) * inc * bin4  + 
           (b5i + u5i) * inc * bin5 + (b6i + u6i) * inc * bin6) %>% 
  # calculate inverse cloglog
  mutate(prob = clogloglink(cloglog, inverse = TRUE)) %>%  
  # calculate event
  mutate(event = rbinom(n(), 1, prob)) 
head(d1)
str(d1)
summary(d1)
glimpse(d1)

## save initial data
# write_csv(d1, "Tutorial_4_planning/data/d1.csv") #
```

summarise the data produced so far

```{r summarise-sim-d1}
d1_pid <- d1 %>%
  group_by(subj, condition, bin) %>%
  summarise(
    across(cloglog:event, \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
    )
d1_pid

d1_group <- d1 %>%
  group_by(condition, bin) %>%
  summarise(
    across(cloglog:event, \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
    )
d1_group
```

a quick plot

cloglog

```{r plot-d1-clog}
p5.1 <- ggplot(d1_group, aes(x=bin, y = cloglog, colour = condition)) +
   geom_line(aes(group = condition)) + 
   geom_point() +
   geom_jitter(data=d1_pid, alpha = 0.5, width = 0.1, height = 0) +
   # scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
   scale_x_continuous(breaks = 1:bin_n, labels = 1:bin_n) +
   scale_fill_brewer(palette = "Dark2") +
   scale_colour_brewer(palette = "Dark2") +
   labs(title = "cloglog - simulation d1 (N=10)",
        x = "time bins")
p5.1
```

event prob

```{r plot-d1-prob}
p5.2 <- ggplot(d1_group, aes(x=bin, y = event, colour = condition)) +
   geom_line(aes(group = condition)) + 
   geom_point() +
   geom_jitter(data=d1_pid, alpha = 0.5, width = 0.1, height = 0) +
   scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
   scale_x_continuous(breaks = 1:bin_n, labels = 1:bin_n) +
   scale_fill_brewer(palette = "Dark2") +
   scale_colour_brewer(palette = "Dark2") +
   labs(title = "event prob - simulation d1 (N=10)",
        x = "time bins")
p5.2
```

convert the data into outcome data (this next chunk can be incorporated into the data structure chunk above, I'm just doing separately here, so that you can see what is initially generated)

```{r calculate-outcome-d1}
d1_out <- d1 %>% 
  group_by(subj, rep, condition) %>%
  mutate(cumsum = cumsum(event),
         outcomel = event==1 & cumsum(event) < 2,
         outcomen = if_else(event==1 & cumsum(event) == 1, 1,
                     if_else(cumsum(event) >= 1, NA, 0))) %>% 
  drop_na(outcomen) %>% 
  ungroup()
head(d1_out)

## save outcome data
# write_csv(d1_out, "Tutorial_4_planning/data/d1_out.csv")
```

summarise the outcome data, which is produced based on the prob value, right?

```{r summary-d1-out}
d1_out_pid <- d1_out %>% 
  group_by(subj, condition, bin) %>% 
  summarise(outcome = mean(outcomen)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_pid

d1_out_group <- d1_out %>% 
  group_by(condition, bin) %>% 
  summarise(outcome = mean(outcomen)) %>% 
  mutate(prime = if_else(condition == "cond1", "con", "inc"),
         prime = factor(prime))
d1_out_group
```

plot outcome using the prime variable to make it comparable to the raw data

```{r plot-d1-outcome}
p5.3 <- ggplot(d1_out_group, aes(x = bin, y = outcome, colour = prime)) +
  geom_line(aes(group = prime)) + 
  geom_point() +
  geom_jitter(data=d1_out_pid, alpha=0.5, width = 0.1, height = 0) +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.25)) +
  scale_x_continuous(breaks = 1:bin_n, labels = 1:bin_n) +
  scale_colour_brewer(palette = "Dark2") +
  labs(title = "simulation d1 (N=10)",
       x = "time bins",
       y = "outcome")
p5.3
```

compare the raw data to the simulated data for a single simulated dataset

```{r plot-raw-plus-d1}
p5.4 <- (p3.1 / p5.3) +
   plot_layout(guides = 'collect',
               axes = 'collect')
p5.4

ggsave ("Tutorial_4_planning/figures/raw_vs_d1.jpeg",
        width = 8, height = 8)
```

ok, this looks reasonably sensible.

# 6. Think about effect sizes

take a look at avg hazards and hazard ratios at the group level in the original data

```{r plot-raw}
p3.1
```

this is the descriptive data per bin and condition for hazards
probability of a response/event, given that it has not happened yet

this is the group level summary data plotted in dat_plot

```{r plot-raw-summary}
data_group
```

calculate the hazard ratio

```{r calculate-hazard}
hazard <- data_group %>% 
  pivot_wider(names_from = "prime",
              values_from = "outcome") %>% 
  mutate(diff = con-inc,
         ratio = inc/con)
hazard
